Autonomous Oceanographic Data Collection towards Estimating Excursion Sets of Vector-valued Random Fields

### 2020-06-18 section3_old.tex
Updates from Overleaf

The goal of this section is to present data collection strategies that aim at decreasing the uncertainty on the target excursion set $\es$.

\subsection{Background}
As before, let $\gp$ be a $\no$-variate random field on $\domain$. See Section \ref{sec:bg_and_notation} for a recap of the specific notation used to handle mean and covariance function of multivariate random fields.

In a sequential setting, the point of view is that $n$ data collection steps have already been performed and one wants to choose what data to collect next. After each step, the GRF model is updated using the cokriging equations \ref{eq:cokrig_mean} and \ref{eq:cokrig_cov} and we will denote by $\mathbb{P}_n$ the law of the field conditional on the data gathered up to (and including) stage $n$.

\begin{remark}
The type of data collected at each stage can be of any type (observation of all components of the field at a single location, observation of some components only but at several locations, ...) since the concept of \textit{generalized location} allows the cokriging equations to handle all those case by putting them on equal footage.
\end{remark}

Our goal is then to optimally select the next spatial location $\spatloc_{n+1}$ at which to observe the field (all of its component), where the meaning of optimal is yet to be defined. Note that here again, the concept of \textit{generalized location} allows one to consider settings where only one component of the field can be observed at each step and the startegy has to choose the spatial location as well as the component to observe (heterotopic case). In the same fashion, one could consider settings where several observations are allowed at each stage (batch case).

We will here restrict ourselves to the case where all components of the field are observed at one spatial location at each stage (isotopic case).

@@

@@

The mathematical expression for the optimal design involves a series
of intermixed maximizations over designs and integrals over data. In
practice, the optimal solution is intractable because of the enormous
growth over stages (see e.g. \cite{sucar2015probabilistic} and
\cite{powell2016perspectives}).  Instead, we outline heuristic
strategies.

\textcolor{red}{Here introduce the fact that candidate points are always chosen in a set $\candidates$. Mention that we usually only allow a finite set. Mention that in the specific case of robot planning, we have a waypoint graph and we thus usually take $\candidates$ to be the set of current neighbouring nodes. Mention that yes, $\candidates$ normally depends on the current location (this is for robot, in general setting, depends on the location of the last observations), but we silently suppress this dependency for readability. Nevertheless, one should always remember that the current set of candidates depends on the current setting.}
\subsection{A Naive Sampling Strategy}
\label{naive}

The simplest heuristic for adaptive sampling is to choose the next
sampling location $\spatloc_{n+1}$ based on current excursion probabilities. At stage $\stage$, one reduces the
largest levels of uncertainty by selecting the location $\spatloc_{\stage + 1}$ with
$P_{\stage}(\gp_{\spatloc_{\stage + 1}} \in T)$ closest to $0.5$..

This strategy does not account for the uncertainty in the GP
variables, \textcolor{red}{(C.T.: do we need this sentence? does it mean anything?) but is instead applicable only if one design (node) has EPs
closer to $0.5$ (see Table \ref{tab:sim_rhoab}, row
two).}

Additionally, this strategy does not account for spatial
correlation. This strategy lacks memory of where it has been and where
the uncertainty has been reduced, and therefore susceptible to local
minima.

@@

Based on the currently available data the myopic strategy selects the location that leads to the biggest expected reduction in IBV:
\begin{criterion}[Myopic]
\begin{equation}\label{critSEQ}
    \spatloc_{\stage + 1} = \argmin_{\spatloc \in \candidates} \EIBV_{\stage}\left(\spatloc\right)
\end{equation}
\end{criterion}

The expected IBV might be efficiently computed for each candidate points using Proposition \ref{propo2}. Once the best location has been selected, a stage of observation is performed there, the GP model is updated, yielding a conditional law $\mathbb{P}_{\stage + 1}$ and the process is repeated.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it can be implemented in a way that is not too demanding on computational power, making it well-suited for embedding on autonomous mobile survey platform (see Section \ref{sec:case_study} for an example).

\subsection{Look-ahead Trajectory Planning}\label{sec:LA}

We now extend the myopic strategy to a look-ahead strategy. We consider two stages of
measurements, and the strategy is then optimal in the sense that it accounts consistently for the expectations and minimizations in these two stages, but still without including any planning beyond those two steps.
\midskip
Say we have collected data up to stage $n$. The principle of two-steps lookahead
strategies is to select the next observation location $\spatloc_{\stage + 1}$ as follows

  \begin{itemize}
      \item First, do as if the $\stage + 1$ observation had already been done and say
          we observed $y_{\stage + 1}$ at $\spatloc_{\stage + 1}$.
      \item If we were at this stage, then we would choose the next points by
          minimizing the expected IBV under the current ($\stage + 1$) conditional
          distribution
          \[
              \spatloc_{\stage + 2} = \argmin_{\spatloc \in \candidates}~ \EIBV_{\stage + 1}\left(\spatloc \mid \spatloc_{\stage + 1},
          y_{\spatloc + 1}\right)
          \]
      \item Now the above expression depends on the observed data $y_{\stage + 1}$,
          whose value we don't know beforehand. What we know is that
          at step $\stage$ it is distributed according to the conditional distribution
          $\mathbb{P}_{\stage}$. We can thus replace it by a random variable $Y$
          following the current conditional law and then average over it.
  \end{itemize}

  Let us now try to make the above more precise.
  \begin{criterion}[2-steps lookahead]
      The next observation location $\spatloc_{\stage + 1}$ is selected according to
      \begin{align}\label{critLA}
          \spatloc_{\stage + 1} = \argmin_{\spatloc \in \candidates}~ \mathbb{E}_{\stage}\left[\min_{\spatloc' \in
                  \candidates(\spatloc)} \EIBV_{\stage + 1}\left(\spatloc' \mid \spatloc,
      Y\right)\right]
      \end{align}
  where $Y$ is a random variable distributed according to the conditional
  law of $\gp_{\spatloc}$ at step $\stage$ (note it silently depends on the considered point
  $\spatloc$) and $\EIBV_{\stage +  1}(\spatloc'\mid \spatloc,Y)$ is the
  expected IBV with the expectation taken according to the conditional law $\mathbb{P}_{\stage}(\cdot\mid
  \spatloc, Y)$. Note that the dependence of the set of candidates on the current location has been made explicit for the second stage of measurements.
  \end{criterion}

\begin{remark}
In a practical setting, the first expectation might be computed by Monte
Carlo sampling of data $Y$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV provided b< Proposition \ref{propo2}, now with conditioning to the first stage data already going into the mean and covariances in Eq. \eqref{eq:krigSUR}. 
\end{remark}

>>>>

An important concept underlying this work is the potential of robotic
sampling to adapt survey plans based on data and statistical methods
in accordance with the sense-plan-act control loop
(Fig. \ref{fig:envir1}). We now present a few adaptive strategies for
robotic sampling that build on the notion of EIBV.

\subsection{Background}

@@

Our adaptive AUV survey is split into many stages. The selection on
where to sample is constrained by a triangular waypoint graph and
defined by its nearest grid nodes.

Let $d^{j,s}$ denote design number $j$ at stage $s$ of a sequential
survey. For the AUV sampling situation, the choice of design entails a commitment to one of many possible navigation directions on the waypoint graph. Once the choice has been made, the AUV steers in the chosen direction and this entails sampling locations $\bu_{d^{j,s}}$ and data $\by_{d^{j,s}}$ is gathered.  We let
$\mathcal{Y}^{s-1} =\{(d^{*,1}, \by_{d^{*,1}}),\ldots, (d^{*,s-1},
\by_{d^{*,s-1}}) \}$ denote the set of all data gathered until stage
$s-1$ according to selected design $d^{*,1},\ldots,d^{*,s-1}$.

The design and data define
the conditional GP with mean
$E(\gp[\x]|\mathcal{Y}^{s-1})$ and (cross) covariance matrix
$\mbox{Var}(\gp[\x]|\mathcal{Y}^{s-1})$. These are efficiently updated in a sequential manner keeping track of the mean and covariance kernel of the GP over the stages. As extensions of Eq. \eqref{eq:cokrig} and \eqref{eq:krig:cov}, we have
\begin{eqnarray}\label{eq:krigSUR}
\mu^{s}(\x) &=&\mu^{s-1}(\x)+\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T (\by_{d^{j,s}}-G_{d^{j,s}} \mu(\x)), \nonumber \\
k^{s}(\x,\x') &=& k^{s-1}(\x,\x')-\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T K_{Y_{d^{j,s}}} \lambda^{[s,s]}(\x',\bu_{d^{j,s}}),
\end{eqnarray}
where $\lambda^{[s,s]}(\x,\bu_{d^{j,s}})=K^{-1}_{Y_{d^{j,s}}} G_{d^{j,s}} k(\bu_{d^{j,s}},\x)$ and the data covariance is $K^{-1}_{Y_{d^{j,s}}}=G_{d^{j,s}} K^s(\bu_{d^{j,s}},\bu_{d^{j,s}})G^T_{d^{j,s}}+R_{d^{j,s}}$, with measurement model as in Eq. \eqref{eq:meas}.

@@

%The optimal path
%selection situation is depicted in Fig. %\ref{fig:PathSelOpt},
%\begin{figure}[b!]
%\centering
%\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
%\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
%\end{figure}
%where design choices are indicated by squares while %data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design involves a series
of intermixed maximizations over designs and integrals over data. In
practice, the optimal solution is intractable because of the enormous
growth over stages (see e.g. \cite{sucar2015probabilistic} and
\cite{powell2016perspectives}).  Instead, we outline heuristic
strategies.

%The derived closed-form solutions for the EIBV are still important building blocks when constructing the sampling designs, as they will be used to score the different adaptive designs. % Efficient
% calculation is important for using adaptive survey designs with
% robotic platforms.

\subsection{A Naive Sampling Strategy}
\label{naive}

The simplest heuristic for adaptive sampling is to choose the next
sampling location based on current probabilities. One reduces the
largest levels of uncertainty by selecting the design $j$ with
$P(Z(\x_j) \in T)$ closest to $0.5$.  With temperature and salinity
variables, this {\it{naive}} strategy selects the stage $s$ sampling
design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j} |P(Z_1(\x_j) \leq t_1, Z_2(\x_j) \leq t_2 | \mathcal{Y}^{s-1})-0.5|, \nonumber
\end{eqnarray}
where $j$ is indexing all neighboring locations on the waypoint graph.

This strategy does not account for the uncertainty in the GP
variables, but is instead applicable only if one design (node) has EPs
closer to $0.5$ (see Table \ref{tab:sim_rhoab}, row
two). Additionally, this strategy does not account for spatial
correlation. This strategy lacks memory of where it has been and where
the uncertainty has been reduced, and therefore susceptible to local
minima.

@@

Based on the currently available data $\mathcal{Y}^{s-1}$, we fit an
updated GP model. Focusing on the EIBV criterion, at the next stage
$s$, the selected design is:
\begin{eqnarray}\label{critSEQ}
    d^{*,s} &=& \mbox{argmin}_{j} \left\{ \int_{\mathcal{M}} E_{\by_{d^{j,s}}|\mathcal{Y}^{s-1}} \left( p(\x;\mathcal{Y}^{j,s})\left( 1-p(\x;\mathcal{Y}^{j,s})\right) \right) d\mes(u), \right\} \\
    p(\x;\mathcal{Y}^{j,s}) &=& P(Z_u \in T |\by_{d^{j,s}},\mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}

Note that this strategy gives a sequential conditional version of
expression (\ref{eq:eibv}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is conditional to this data. A similar closed-form
calculation for EIBV is hence applicable in Eq.  \eqref{critSEQ} using the updated
GP model from stage $s-1$ as defined via Eq. \eqref{eq:krigSUR} in the results from Proposition \ref{propo2}. Once the data are collected for the best design $d^{*s}$ at stage $s$, the GP model is updated again. The mean and covariance are used to compute the next design at stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it is easily implemented, in our case, onboard
an AUV, using an efficient data update of the GP model and the
calculation of the closed-form EIBV expressions for each subsequent
trajectory.

\subsection{Look-ahead Trajectory Planning}
\label{sec:LA}

We now extend the myopic strategy to a look-ahead strategy. We consider two stages of
measurements, and the strategy is then optimal in the sense that it accounts consistently for the expectations and minimizations in these two stages, but still without including any planning beyond those two steps. In addition to the
next stage of measurements $\by^{j,s}$, this look-ahead strategy anticipates the subsequent design indexed $j_2$ with data
$\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$. The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j} \left\{ E_{\by_{d^{j,s}}|\mathcal{Y}^{s-1}} \left( \mbox{argmin}_{j_2}  V_{j_2} \right) \right\}, \\
V_{j_2} & = & \int_{\mathcal{M}} E_{\by_{d^{j_2,s+1}}|\mathcal{Y}^{j,s}} \left( p(\x;\mathcal{Y}^{j,s+1}) ( 1-p(\x;\mathcal{Y}^{j,s+1}) ) \right) d\mes(u), \nonumber \\
    p(\x;\mathcal{Y}^{j,s+1}) &=& P(Z(\x) \in T |\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by_{d^{j,s}},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}\}$
represent the sets of data variables, and the expectations are
calculated with respect to the conditional densities given data at
that stage.  We solve the first expectation by Monte
Carlo sampling of data $\by_{d^{j,s}}$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV using the results from Proposition \ref{propo2}, now with conditioning to the first stage data already going into the mean and covariances in Eq. \eqref{eq:krigSUR}.

Even though the strategy looks at two sampling stages, it is only used to find
the current best design. When data are collected at stage $s$, the GP model is
updated as in Eq. \eqref{eq:krigSUR}, and the mean and (cross) covariance are
used to compute the next design for stage $s+1$, now anticipating what
stage $s+1$ and $s+2$ could offer, and so on.
This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation onboard the AUV this is reaching the limits allowed for calculation. We conduct some pruning of the least valuable design paths in the evaluation of Eq. \eqref{critLA}, to get a satisfactory computation time below one minute per waypoint. This means that we use the myopic
strategy to rank the three best designs following the first stage alone, and
for each of these preferred designs, we undertake the look-ahead calculations.
